---
title: "Week 7 Lab: Introduction to R's Data Structures"
subtitle: "Student Template (Completed)"
author: "Jane Doe"
format: 
  html:
    theme: cosmo
    toc: true
---

## 1. Introduction

This week in lecture, we discussed **Chapter 7: Sampling**. In R, the "subset" we select from a population is stored in a **data frame**. Today, we will learn how to access data from our course package, view it, and identify its fundamental components: observations and variables.

## 2. Setup

In the code chunk below, load the two packages we need for today's lab: `tidyverse` and `mccoursepack`.

```{r}
#| label: setup
#| message: false

# TODO: Load the tidyverse and mccoursepack packages
library(tidyverse)
library(mccoursepack)
```

## 3. Accessing Our Sample Data

The datasets for this course are included in the `mccoursepack` package. This means after loading the package, the data is available to use by name.

In the chunk below, type the name of the `simpsons_characters` dataset to see a preview of it.

```{r}
#| label: access-data

# TODO: Type the name of the simpsons_characters object here to preview it
simpsons_characters
```

## 4. Viewing the Data Frame

The `View()` function (with a capital "V") opens your data in a new tab, which is very useful for exploring.

In the chunk below, use the `View()` function to open the `simpsons_characters` dataset.

```{r}
#| label: view-data
#| eval: false

# TODO: Use the View() function on the simpsons_characters dataset
View(simpsons_characters)
```

**After viewing the data, answer these questions:**

1.  What is the **unit of observation** in this dataset? (i.e., what does each row represent?)
    
    > Each row represents a single, unique character from The Simpsons.
2.  How many observations are in this sample? How many variables?
    
    > There are 6,722 observations (characters) and 4 variables.

## 5. Manually Creating a Data Frame

We can also create a data frame from scratch using the `tibble()` function. In the chunk below, write the code to create a data frame named `simpson_family` that looks like this:

| name | gender | age |
| :--- | :--- | --: |
| Homer Simpson | m | 39 |
| Marge Simpson | f | 36 |
| Bart Simpson | m | 10 |
| Lisa Simpson | f | 8 |
| Maggie Simpson | f | 1 |

```{r}
#| label: create-tibble

# TODO: Create the simpson_family tibble here
simpson_family <- tibble(
  name = c("Homer Simpson", "Marge Simpson", "Bart Simpson", "Lisa Simpson", "Maggie Simpson"),
  gender = c("m", "f", "m", "f", "f"),
  age = c(39, 36, 10, 8, 1)
)

simpson_family
```

## 6. Conclusion

Congratulations on completing your first R lab! You've successfully loaded packages, accessed and viewed data, and learned the difference between an observation and a variable.
