---
# -----------------------------------------------------------------
# APA 7TH MANUSCRIPT: QUARTO TEMPLATE
# -----------------------------------------------------------------
# This YAML header controls the formatting of your entire document.
# It is pre-configured for a professional APA 7th Edition paper.
title: "Your Full Research Paper Title Goes Here"
author:
  - name: "Your Full Name"
    affiliation: "Your University"
    author-note: "first-author-note" # This links to the Author Note section below

# ---------------------------------------------------------------
# PROJECT STARTER INSTRUCTIONS (Read Me First)
# ---------------------------------------------------------------
# HOW TO USE THIS TEMPLATE:
# 1. Fill in the YAML header below with your project details.
# 2. Write your manuscript in the sections provided.
# 3. Update the bibliography file (references.bib) as you add sources.
# 4. Replace the dummy data in /data/survey_data.csv with your real survey data.
# 5. Use RStudio's 'Render' button or run `quarto render research_paper.qmd` to create your PDF in /output.
# 6. Review the checklist below to ensure you meet all requirements.
# 7. For technical help, see:
#    - grad_technical_guide.md
#    - grad_github_guide.md
#    - grad_zotero_guide.md
#    - Quarto docs: https://quarto.org/docs/
#
# MASTER PROJECT CHECKLIST (Quick Reference)
# - [ ] 6,000–8,000 words (excluding references, tables, appendices)
# - [ ] 15-20 academic, peer-reviewed citations
# - [ ] Two Research Questions (RQs) and/or Hypotheses (RHs)
# - [ ] All data analyzed in RStudio
# - [ ] At least two APA-formatted data visualizations
# - [ ] APA 7th Edition style throughout
# - [ ] Final PDF rendered and saved in /output
# - [ ] All files pushed to GitHub
# - [ ] PDF submitted to Blackboard
#
# For more details, see the README.md and your course guides.
    orcid: "0000-0000-0000-0000"     # Get your free ORCID iD at orcid.org

# Add more authors by uncommenting the lines below
#  - name: "Second Author's Name"
#    affiliation: "Second Author's University"
#    author-note: "second-author-note"
#    orcid: "0000-0000-0000-0000"

# --- Author Notes ---
# 
author-note-place: "margin" # Puts notes in the footer (common) or "margin"
author-note:
  - id: "first-author-note"
    text: |
      **Disclosures:** [If you have no conflicts of interest, state: "We have no known conflict of interest to disclose.". If you received funding, state it here.]

      **Correspondence:** [Correspondence concerning this article should be addressed to Your Full Name, Department Name, University Name, University Address. Email: your.email@university.edu ]

#  - id: "second-author-note"
#    text: "Department of [Name], [University Name]."

# --- Abstract & Keywords ---
abstract: |
  [Your abstract is a brief, comprehensive summary of the paper. It should be no more than 250 words. Summarize your problem, participants, method, basic findings, and conclusions. This is often the last part you write.]

keywords: "keyword 1, keyword 2, keyword 3, keyword 4, keyword 5" # [Provide 3-5 relevant keywords ]

# --- APA & Document Settings ---
running-head: "YOUR SHORTENED TITLE" # [Must be 50 characters or less, in ALL CAPS ]
bibliography: references.bib         # [File name for your Zotero .bib file ]
format: 
  apa-pdf:
    keep-tex: false
    watermark: false # Set to true for a DRAFT watermark
    
# --- R Code Setup ---
# This global chunk sets default options for all R code blocks
execute:
  echo: false     # Hides the R code itself in the final PDF
  warning: false  # Hides any R warnings
  message: false  # Hides any R messages
---

```{r}
#| label: setup
#| include: false
#| 
# Load required R packages
library(tidyverse) # For data manipulation (dplyr, ggplot2) and reading (readr)
library(here)      # For locating files in your RStudio Project
library(knitr)     # For creating formatted tables
````

## Literature Review

[Your literature review text goes here. Synthesize your 15-20 sources to establish the background and importance of your topic. Identify the gap in the research that your study will address.]

## The Present Study

[This section narrows the focus from the general lit review to your specific project. End this section by clearly stating your two Research Questions and/or Hypotheses.]

**RQ1:** [State your first research question here.]
**RQ2:** [State your second research question here.]

# Method

## Participants

[Describe your participants. How many? (N=xxx). Provide key demographics (age, gender, etc.). [cite\_start]See Table 7.2  for a good example.]

[cite\_start][Describe your sampling procedure (e.g., convenience, snowball) and how you recruited (e.g., social media, SONA).]

## Measures

[Describe the survey instrument you created. If you used established scales, describe them (e.g., "News trust was measured using a 5-item Likert scale...").]

## Procedure

[Explain the steps you took. "Data was collected via an online survey built in [Qualtrics/Google Forms] and distributed from [Date] to [Date]. Participants first read an informed consent page..."]

## Analytic Strategy

[Briefly state how you will analyze the data to answer your RQs. (e.g., "To answer RQ1, a Pearson correlation was computed... To answer RQ2, an independent-samples t-test was conducted...")]

# Results

```{r}
#| label: load-data

# Load data using the 'here' package to find the file
file_path <- here("data", "survey_data.csv")
survey_data <- read_csv(file_path)

# --- Data Cleaning (Example) ---
# This is where you would clean your data.
# e.g., filter(age > 18), mutate(gender = as.factor(gender)), etc.
cleaned_data <- survey_data
```

[Start with a brief overview of your analysis. Then, address each RQ/RH in order.]

## Research Question 1

[State the results for your first RQ/RH. Include the relevant statistics in the text. [cite\_start]For example: "A Pearson correlation revealed a significant negative relationship between social media use and news trust, *r*(98) = -.25, *p* = .012." ]

```{r}
#| label: tbl-corr-matrix
#| tbl-cap: "Correlation Matrix for Key Study Variables"
#| tbl-note: "N = 100. *p* < .05, **p* < .01."
#| tbl-cap-location: top # APA format (Quarto default is bottom)

# --- Dummy Code for a Correlation Table ---
# This code won't run without a 'cor.R' file, 
# but it shows the student the *structure* of making a table.
# A simpler way:
corr_matrix <- cleaned_data |>
  select(age, social_media_use, news_trust, political_knowledge) |>
  cor()

# Use knitr::kable() to format the table for APA
kable(corr_matrix, 
      digits = 2, 
      booktabs = TRUE,
      caption = "Correlation Matrix") |>
  # This is an example of a kableExtra function to add a footnote
  # kable_classic(full_width = F) |>
  # add_footnote("N = 100. *p* < .05", notation = "symbol")
```

[Your text continues here. You must refer to the table above, e.g., "Full results of the correlation analysis are presented in @tbl-corr-matrix."]

## Research Question 2

[State the results for your second RQ/RH. Include relevant statistics in the text.]

```{r}
#| label: fig-bar-chart
#| fig-cap: "Political Knowledge Scores by Gender."
#| fig-cap-location: top # APA 7th (Quarto's default is bottom)
#| fig-alt: "A bar chart showing that mean political knowledge was higher for participants identifying as Male than for those identifying as Female."
#| 
#| # --- Figure Note ---
#| # Quarto's `apa-pdf` format doesn't have a `fig-note` chunk option
#| # like it does for tables (`tbl-note`).
#| # Instead, you add the Figure Note manually in the text block
#| # immediately AFTER the figure chunk.

# --- Dummy Code for a ggplot2 Figure ---
cleaned_data |>
  filter(gender %in% c("Male", "Female")) |> # Example filter
  ggplot(aes(x = gender, y = political_knowledge, fill = gender)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  geom_errorbar(stat = "summary", fun.data = "mean_cl_normal", width = 0.2) +
  labs(y = "Political Knowledge Score (0-10)", x = "Gender") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none") # Remove legend as x-axis is clear
```

*Note.* This figure displays the mean political knowledge scores for male and female participants. [cite\_start]Error bars represent 95% confidence intervals.

[Your text continues here. You must refer to the figure above, e.g., "As shown in @fig-bar-chart, male participants (M = 7.2, SD = 1.1) scored significantly higher..."]

# Discussion

[Summarize your main findings in plain English. Do not just repeat the statistics from the Results section.]

[Discuss how your findings support or challenge existing literature. Connect back to the sources from your introduction.]

## Limitations

[Acknowledge the limitations of your study (e.g., small/non-representative sample, reliance on self-report measures).]

## Future Directions

[Suggest new research questions that arise from your findings. What should the next study in this area investigate?]

## Conclusion

[Provide a final, summary paragraph. What is the key takeaway message of your research? What are the practical implications of your work?]

# References

::: {\#refs}
:::

# Appendix A: Survey Instrument

[You can paste the text of your survey questions here.]

````