Here is the updated Markdown file for the `janitor` package.

I've gone through and updated all the R code examples to use the new, correct column names from your `w144_wrangled` data (e.g., `QKEY`, `age_group`, `gender`, `party_simple`, `education`). I've also updated the filter conditions, as the new text-based columns use `NA` for missing values instead of the number `99`.

-----

# R `janitor` Package: A Guide for Students

This guide introduces you to the most common functions from the **`janitor`** package. This package is designed to make data cleaning and creating summary tables (like crosstabs) simple and efficient. It works perfectly with the `dplyr` verbs you've already learned.

All examples assume you have first loaded the `dplyr` and `janitor` libraries and the Pew dataset:

```{r}
# First, make sure you have janitor installed: install.packages("janitor")
library(dplyr)
library(janitor) # Load janitor
library(readr)

# Load the dataset
# This R code below loads the *original* data and *creates*
# the 'w144_wrangled' data frame by adding new columns.
data_path <- system.file("extdata", "ATP_W144_excerpt.csv", package = "mccoursepack")
pew_data <- read_csv(data_path)

pew_wrangled <- pew_data %>%
  mutate(
    party_simple = case_when(
      F_PARTY_FINAL == "Rep/lean Rep" ~ "Republican",
      F_PARTY_FINAL == "Dem/lean Dem" ~ "Democrat",
      TRUE ~ "Independent/Other"
    ) %>% factor(levels = c("Republican", "Democrat", "Independent/Other")),
    age_cat = as_factor(F_AGECAT),
    education = as_factor(F_EDUCCAT),
    gender = as_factor(GENDER),
    income = as_factor(F_INC_SDT1)
  )
```

-----

## 🧹 Core Cleaning Functions

These functions help you clean your dataset.

### `clean_names()`

**Explanation:** This is the most famous `janitor` function. It "perfectly formats data.frame column names." It converts all names to a consistent format (by default, "snake_case," which is `lowercase_and_with_underscores`) and ensures all names are unique. This is extremely useful when you import data from Excel or other sources with messy column names (like "Column Name (2025)").

**Example:** Let's create a small, "messy" data frame and clean it.

```{r}

# Create a messy data frame using our pew_wrangled
# We select real columns and then rename them to be messy
messy_data <- pew_wrangled %>%
  select(QKEY, education, age_cat) %>%
  rename("Respondent ID" = QKEY,
         "Level of Education" = education,
         "age_cat" = age_cat) # This one is already mostly clean

print("--- Original Messy Names ---")
glimpse(messy_data)

# Now, clean the names
clean_data <- messy_data %>%
  clean_names()

print("--- Cleaned Names ---")
glimpse(clean_data)
```

**Explanation:** We first `rename()` our columns to have spaces and capital letters, which are difficult to work with in R. When we pipe `messy_data` into `clean_names()`, it automatically converts "Respondent ID" to `respondent_id` and "Level of Education" to `level_of_education`. The `age_group` column is also standardized.

**Practice Prompt:** The `pew_wrangled` object is already clean. To see `clean_names()` in action, pipe the original `pew_data` object into `clean_names()` and use `glimpse()` to see the result.

-----

### `remove_empty()`

**Explanation:** This is a simple utility to remove all rows or all columns from your data that are *entirely* `NA`. You can specify `which = "rows"`, `which = "cols"`, or `which = c("rows", "cols")`.

**Example:** Let's create a test data frame with an empty row and an empty column.

```{r}
# Create a test data frame
test_df <- data.frame(
  A = c(1, NA, 3),
  B = c(NA, NA, NA), # An all-NA column
  C = c(4, 5, NA)
)
# Add an all-NA row
test_df[4, ] <- NA

print("--- Original Data ---")
print(test_df)

# Remove both empty rows and columns
clean_df <- test_df %>%
  remove_empty(which = c("rows", "cols"))

print("--- Cleaned Data ---")
print(clean_df)
```

**Explanation:** The original `test_df` had an all-NA column (`B`) and an all-NA row (row 4). The `remove_empty()` function identified and removed both, returning only the 3x2 data frame with actual data.

**Practice Prompt:** Using the same `test_df`, how would you remove *only* the empty columns but *keep* the empty row?

-----

### `get_dupes()`

**Explanation:** This function helps you find duplicate records during data cleaning. [cite: 712] You provide the data frame and the column names to check for duplicate combinations. It returns a new data frame showing *only* the rows that are duplicates, plus a new column `dupe_count` showing how many times that combination appeared.

**Example:** Let's check if there are any duplicate respondent IDs (`QKEY`) in our dataset. (There shouldn't be!)

```{r}

# Check for duplicates in the 'QKEY' column
pew_wrangled %>%
  get_dupes(QKEY)
```

**Explanation:** This code checks the entire `pew_wrangled` data frame to see if any value in the `QKEY` column appears more than once. If it returns an empty data frame (with 0 rows), it means our data is "clean" and every respondent ID is unique.

**Practice Prompt:** Check if there are any rows that have the *exact same combination* of `age_cat`, `gender`, and `party_simple`.

-----

## 📊 Tabulation Functions (`tabyl`)

This family of functions is for creating frequency tables and crosstabs.

### `tabyl()`

**Explanation:** This is a modern replacement for R's `table()` function. It takes a data frame and one, two, or three variable names and creates a frequency table.

  - **One variable:** Creates a frequency table with counts (`n`) and percentages (`percent`).
  - **Two variables:** Creates a crosstabulation (crosstab) showing the counts of co-occurrences.
  - The result is a **data frame**, so you can pipe it to other `dplyr` or `janitor` functions.

**Example:** Create a 1-way frequency table for education level and a 2-way crosstab for gender and political party.

```{r}

# 1-way table (using education)
pew_wrangled %>%
  tabyl(education)

# 2-way table (crosstab)
# We can filter out NA values for a cleaner table
pew_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple)
```

**Explanation:** The first `tabyl()` shows the counts and percentages for `education`. The second `tabyl()` creates a table with `gender` as the rows and `party_simple` as the columns, showing the count of respondents at each intersection (e.g., how many people are both "Female" and "Democrat").

**Practice Prompt:** Create a two-way `tabyl()` to explore the relationship between `age_cat` and `party_simple`.

-----

### `adorn_totals()`

**Explanation:** This function is "piped" after a `tabyl()` to add a totals row and/or column. You can specify `where = "row"`, `where = "col"`, or `where = c("row", "col")` to control which totals are added.

**Example:** Create a crosstab of gender and party and add totals for both rows and columns.

```{r}
pew_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_totals(where = c("row", "col"))
```

**Explanation:** This chain first creates the crosstab. Then, `adorn_totals()` adds a "Total" row at the bottom and a "Total" column on the right, giving you the marginal frequencies.

**Practice Prompt:** Create a `tabyl()` of `age_group` and `gender` (filtering for `NA`s) and add *only* a totals column.

-----

### `adorn_percentages()`

**Explanation:** This function converts a data frame of counts (from `tabyl()`) into percentages. You **must** specify the `denominator` to tell it *how* to calculate the percentages:

  - `denominator = "row"` (calculates across the rows)
  - `denominator = "col"` (calculates down the columns)
  - `denominator = "all"` (calculates based on the grand total)

**Example:** Answer the question: "What percentage of *each political party* is "Female" vs. "Male"?"

```{r}
# We must use denominator = "col" to get percentages *within* each column (party)
pew_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_percentages(denominator = "col")
```

**Explanation:** This calculates the percentages *down the columns*. For example, you'll see the percentage of "Democrat" respondents who are "Female" and the percentage who are "Male". Each column will add up to 1.0 (representing 100% of that party).

**Practice Prompt:** How would you answer the *other* question: "What percentage of *each gender* falls into each political party?" (Hint: You'll need to change the `denominator`).

-----

### `adorn_pct_formatting()`

**Explanation:** This function makes your percentages (which are decimals, like `0.5833`) look pretty. It multiplies by 100, rounds the number, and adds a "%" sign. You can control the number of decimal places with the `digits` argument.

**Example:** Create a column-percentage table and format it to one decimal place.

```{r}
pew_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 1)
```

**Explanation:** This chain builds on the last example. The `adorn_percentages()` output (e.g., `0.5833...`) is piped into `adorn_pct_formatting()`, which turns it into a clean, readable string like "58.3%".

**Practice Prompt:** Create the same table, but show the percentages with *zero* decimal places.

-----

### `adorn_ns()`

**Explanation:** This is the final piece for a "publication-ready" table. After you've converted your counts to percentages, `adorn_ns()` adds the original counts (the "N") back into the table. This allows you to show both the percentage and the N in a single cell (e.g., "58.3% (120)").

**Example:** Create a full table showing formatted column percentages *and* the underlying Ns.

```{r}
pew_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_totals("col") %>% # Add column totals before percentages
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns(position = "front")
```

**Explanation:** This is a common "full" pipeline.

1.  `tabyl()`: Creates the counts.
2.  `adorn_totals("col")`: Adds a "Total" column.
3.  `adorn_percentages("col")`: Calculates column percentages (this works correctly even with the total column!).
4.  `adorn_pct_formatting()`: Formats the percentages to strings (e.g., "58.3%").
5.  `adorn_ns(position = "front")`: Adds the original Ns (e.g., "(120)") to the front of the percentage string.

**Practice Prompt:** Create the same table, but pipe `adorn_totals(where = c("row", "col"))` *after* `adorn_ns()` to get a fully-adorned table with Ns, percentages, and totals for both dimensions.
Here is the updated Markdown file for the `janitor` package.

I've gone through and updated all the R code examples to use the new, correct column names from your `w144_wrangled` data (e.g., `QKEY`, `age_group`, `gender`, `party_simple`, `education`). I've also updated the filter conditions, as the new text-based columns use `NA` for missing values instead of the number `99`.

-----

# R `janitor` Package: A Guide for Students

This guide introduces you to the most common functions from the **`janitor`** package. This package is designed to make data cleaning and creating summary tables (like crosstabs) simple and efficient. It works perfectly with the `dplyr` verbs you've already learned.

All examples assume you have first loaded the `dplyr` and `janitor` libraries and the Pew dataset:

```{r}
# First, make sure you have janitor installed: install.packages("janitor")
library(dplyr)
library(janitor) # Load janitor
library(readr)

# Load the dataset
# This R code below loads the *original* data and *creates*
# the 'w144_wrangled' data frame by adding new columns.
data_path <- system.file("extdata", "w144_teaching_dataset_v2.csv", package = "mccoursepack")
w144_data <- read_csv(data_path)

w144_wrangled <- w144_data %>% 
  mutate(
    party_simple = case_when(
      party == "Republican" ~ "Republican",
      party == "Democrat" ~ "Democrat",
      TRUE ~ "Independent/Other"
    ) %>% 
      factor(levels = c("Republican", "Democrat", "Independent/Other")),
    
    platform_count = (uses_facebook == "Yes") +
                     (uses_x == "Yes") +
                     (uses_instagram == "Yes") +
                     (uses_tiktok == "Yes"),
    across(
      starts_with("fb_why_"), 
      ~ if_else(. == "Yes", 1, 0),
      .names = "{.col}_n" 
    )
  ) %>%
  mutate(
    fb_uses_count = rowSums(
      across(starts_with("fb_why_") & ends_with("_n")), 
      na.rm = TRUE
    )
  )
```

-----

## 🧹 Core Cleaning Functions

These functions help you clean your dataset.

### `clean_names()`

**Explanation:** This is the most famous `janitor` function. It "perfectly formats data.frame column names." It converts all names to a consistent format (by default, "snake\_case," which is `lowercase_and_with_underscores`) and ensures all names are unique. This is extremely useful when you import data from Excel or other sources with messy column names (like "Column Name (2025)").

**Example:** Let's create a small, "messy" data frame and clean it.

```{r}
# Create a messy data frame using our w144_wrangled
# We select real columns and then rename them to be messy
messy_data <- w144_wrangled %>%
  select(QKEY, education, age_group) %>%
  rename("Respondent ID" = QKEY,
         "Level of Education" = education,
         "age_group" = age_group) # This one is already mostly clean

print("--- Original Messy Names ---")
glimpse(messy_data)

# Now, clean the names
clean_data <- messy_data %>%
  clean_names()

print("--- Cleaned Names ---")
glimpse(clean_data)
```

**Explanation:** We first `rename()` our columns to have spaces and capital letters, which are difficult to work with in R. When we pipe `messy_data` into `clean_names()`, it automatically converts "Respondent ID" to `respondent_id` and "Level of Education" to `level_of_education`. The `age_group` column is also standardized.

**Practice Prompt:** The `w144_wrangled` object is already clean. To see `clean_names()` in action, pipe the original `w144_data` object into `clean_names()` and use `glimpse()` to see the result.

-----

### `remove_empty()`

**Explanation:** This is a simple utility to remove all rows or all columns from your data that are *entirely* `NA`. You can specify `which = "rows"`, `which = "cols"`, or `which = c("rows", "cols")`.

**Example:** Let's create a test data frame with an empty row and an empty column.

```{r}
# Create a test data frame
test_df <- data.frame(
  A = c(1, NA, 3),
  B = c(NA, NA, NA), # An all-NA column
  C = c(4, 5, NA)
)
# Add an all-NA row
test_df[4, ] <- NA

print("--- Original Data ---")
print(test_df)

# Remove both empty rows and columns
clean_df <- test_df %>%
  remove_empty(which = c("rows", "cols"))

print("--- Cleaned Data ---")
print(clean_df)
```

**Explanation:** The original `test_df` had an all-NA column (`B`) and an all-NA row (row 4). The `remove_empty()` function identified and removed both, returning only the 3x2 data frame with actual data.

**Practice Prompt:** Using the same `test_df`, how would you remove *only* the empty columns but *keep* the empty row?

-----

### `get_dupes()`

**Explanation:** This function helps you find duplicate records during data cleaning. [cite: 712] You provide the data frame and the column names to check for duplicate combinations. It returns a new data frame showing *only* the rows that are duplicates, plus a new column `dupe_count` showing how many times that combination appeared.

**Example:** Let's check if there are any duplicate respondent IDs (`QKEY`) in our dataset. (There shouldn't be\!)

```{r}
# Check for duplicates in the 'QKEY' column
w144_wrangled %>%
  get_dupes(QKEY)
```

**Explanation:** This code checks the entire `w144_wrangled` data frame to see if any value in the `QKEY` column appears more than once. If it returns an empty data frame (with 0 rows), it means our data is "clean" and every respondent ID is unique.

**Practice Prompt:** Check if there are any rows that have the *exact same combination* of `age_group`, `gender`, and `party_simple`.

-----

## 📊 Tabulation Functions (`tabyl`)

This family of functions is for creating frequency tables and crosstabs.

### `tabyl()`

**Explanation:** This is a modern replacement for R's `table()` function. It takes a data frame and one, two, or three variable names and creates a frequency table.

  - **One variable:** Creates a frequency table with counts (`n`) and percentages (`percent`).
  - **Two variables:** Creates a crosstabulation (crosstab) showing the counts of co-occurrences.
  - The result is a **data frame**, so you can pipe it to other `dplyr` or `janitor` functions.

**Example:** Create a 1-way frequency table for education level and a 2-way crosstab for gender and political party.

```{r}
# 1-way table (using education)
w144_wrangled %>%
  tabyl(education)

# 2-way table (crosstab)
# We can filter out NA values for a cleaner table
w144_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple)
```

**Explanation:** The first `tabyl()` shows the counts and percentages for `education`. The second `tabyl()` creates a table with `gender` as the rows and `party_simple` as the columns, showing the count of respondents at each intersection (e.g., how many people are both "Female" and "Democrat").

**Practice Prompt:** Create a two-way `tabyl()` to explore the relationship between `age_group` and `party_simple`.

-----

### `adorn_totals()`

**Explanation:** This function is "piped" after a `tabyl()` to add a totals row and/or column. You can specify `where = "row"`, `where = "col"`, or `where = c("row", "col")` to control which totals are added.

**Example:** Create a crosstab of gender and party and add totals for both rows and columns.

```{r}
w144_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_totals(where = c("row", "col"))
```

**Explanation:** This chain first creates the crosstab. Then, `adorn_totals()` adds a "Total" row at the bottom and a "Total" column on the right, giving you the marginal frequencies.

**Practice Prompt:** Create a `tabyl()` of `age_group` and `gender` (filtering for `NA`s) and add *only* a totals column.

-----

### `adorn_percentages()`

**Explanation:** This function converts a data frame of counts (from `tabyl()`) into percentages. You **must** specify the `denominator` to tell it *how* to calculate the percentages:

  - `denominator = "row"` (calculates across the rows)
  - `denominator = "col"` (calculates down the columns)
  - `denominator = "all"` (calculates based on the grand total)

**Example:** Answer the question: "What percentage of *each political party* is "Female" vs. "Male"?"

```{r}
# We must use denominator = "col" to get percentages *within* each column (party)
w144_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_percentages(denominator = "col")
```

**Explanation:** This calculates the percentages *down the columns*. For example, you'll see the percentage of "Democrat" respondents who are "Female" and the percentage who are "Male". Each column will add up to 1.0 (representing 100% of that party).

**Practice Prompt:** How would you answer the *other* question: "What percentage of *each gender* falls into each political party?" (Hint: You'll need to change the `denominator`).

-----

### `adorn_pct_formatting()`

**Explanation:** This function makes your percentages (which are decimals, like `0.5833`) look pretty. It multiplies by 100, rounds the number, and adds a "%" sign. You can control the number of decimal places with the `digits` argument.

**Example:** Create a column-percentage table and format it to one decimal place.

```{r}
w144_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 1)
```

**Explanation:** This chain builds on the last example. The `adorn_percentages()` output (e.g., `0.5833...`) is piped into `adorn_pct_formatting()`, which turns it into a clean, readable string like "58.3%".

**Practice Prompt:** Create the same table, but show the percentages with *zero* decimal places.

-----

### `adorn_ns()`

**Explanation:** This is the final piece for a "publication-ready" table. After you've converted your counts to percentages, `adorn_ns()` adds the original counts (the "N") back into the table. This allows you to show both the percentage and the N in a single cell (e.g., "58.3% (120)").

**Example:** Create a full table showing formatted column percentages *and* the underlying Ns.

```{r}
w144_wrangled %>%
  filter(!is.na(gender), !is.na(party_simple)) %>%
  tabyl(gender, party_simple) %>%
  adorn_totals("col") %>% # Add column totals before percentages
  adorn_percentages(denominator = "col") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns(position = "front")
```

**Explanation:** This is a common "full" pipeline.

1.  `tabyl()`: Creates the counts.
2.  `adorn_totals("col")`: Adds a "Total" column.
3.  `adorn_percentages("col")`: Calculates column percentages (this works correctly even with the total column\!).
4.  `adorn_pct_formatting()`: Formats the percentages to strings (e.g., "58.3%").
5.  `adorn_ns(position = "front")`: Adds the original Ns (e.g., "(120)") to the front of the percentage string.

**Practice Prompt:** Create the same table, but pipe `adorn_totals(where = c("row", "col"))` *after* `adorn_ns()` to get a fully-adorned table with Ns, percentages, and totals for both dimensions.