---
title: "Week 7 Lab: From Physical to Digital Sampling"
subtitle: "A Two-Part Lab on Analyzing Samples in R"
author: "Your Name"
format:
  pdf:
    toc: true
    number-sections: true
---

## 1. Welcome & Objectives

Today's lab is a two-part, hands-on activity. First, you will draw and analyze a **physical sample** of dice. Then, you will apply the exact same logic to analyze a **digital sample** from a large dataset containing every line of dialogue from *The Simpsons*. This will show how the core principles of sampling are the same whether you're holding the data in your hand or exploring it on a computer.

**Learning Goals:**

-   Draw a physical sample and enter its data into RStudio.
-   Analyze a digital sample from a large, real-world dataset.
-   See how R can be used to quickly summarize and visualize data from any sample.
-   Understand the connection between a sample and the population it comes from.

## 2. Getting Set Up

First, make sure you are in your RStudio project. Next, load the packages we will use. We are adding `tidytext`, a package for basic text analysis.

### Installing or Updating the Course Package

Before starting the lab, ensure you have the latest version of the `mccoursepack` package installed. Run the following code in your R console:

```{r}
# Install or update the course package
pak::pak("SIM-Lab-SIUE/mccoursepack")
```

### Beginner Note: What does this code do?

When you see `library(...)` in R, it means you are loading a package—a collection of functions and data that help you do specific tasks. You only need to run these lines once at the start of your session.

* `tidyverse` is a set of packages for working with data and making plots.
* `tidytext` helps you analyze text (words, sentences, etc.).
* `mccoursepack` is a special package for this course. It gives you access to the Simpsons data used in this lab. When you load it, datasets like `simpsons_script_lines` become available for you to use in your code.

```{r}
#| label: setup
#| message: false
# Load the necessary packages for this lab
library(tidyverse)      # For data wrangling and visualization
library(tidytext)       # For analyzing text data
library(mccoursepack)   # For the Simpsons dataset
```

## 3. Part 1: Your Physical Sample (Dice 🎲)

Let's start with something tangible. The bag of 118 dice is our **population**.

### Step 1.1: Draw and Tally Your Sample

Go to the bag and, without looking, draw a **simple random sample** of **20 dice**. On a piece of paper, count how many of each die type you have (D4, D6, D8, etc.).

### Step 1.2: Enter Your Data in RStudio

Find the `my_dice_sample` tibble in the code chunk below. **Your only task is to replace the zeros (`0`) in the `count` column with the numbers you just tallied.**

### Beginner Note: What is a tibble?

A `tibble` is a modern version of a data frame in R. It's just a table of data, like a spreadsheet, but inside R. Here, you will enter your sample counts for each type of die.

```{r}
#| label: enter-dice-data
# 👉 YOUR ACTION REQUIRED 👈
# Replace the zeros with your counts. The total should be 20.
my_dice_sample <- tibble(
  die_type = c("D4", "D6", "D8", "D10", "D12", "D20", "D100"),
  count = c(0, 0, 0, 0, 0, 0, 0) # e.g., c(2, 6, 3, 2, 2, 4, 1)
)

# Run this chunk after you've entered your data.
print(my_dice_sample)
```

### Beginner Note: What is ggplot2?

`ggplot2` is a powerful tool for making plots in R. It is part of the `tidyverse` package you loaded earlier. The code below creates a bar chart showing how many of each die type you have in your sample. You don't need to change anything—just run it!

```{r}
#| label: view-dice-plot
# This code creates a bar chart of your dice sample

ggplot(my_dice_sample, aes(x = die_type, y = count, fill = die_type)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "The Composition of My Dice Sample",
    x = "Type of Die",
    y = "Count in Sample"
  )
```

### Step 1.3: Visualize Your Dice Sample

Now for the fun part. The code below will automatically create a bar chart from your specific sample data. Just run the chunk!

### Step 1.4: Reflection

The true proportion of D6s in the bag is **28.8%**. The difference between your sample's result and this true value is called **sampling error**. It's the natural, unavoidable variation that comes from studying a sample instead of the whole population.

## 4. Part 2: Your Digital Sample (The Simpsons 📺)

Great! You've analyzed a physical sample. Now, let's apply the exact same logic to a much larger population: all \~158,000 lines of dialogue in the `simpsons_script_lines` dataset. Our goal is to see what a sample can tell us about a character.

### Step 2.1: "Draw" Your Digital Sample

We can't put 158,000 lines of text in a bag, but we can use code to do the same thing. The code below takes a **simple random sample** of 100 lines spoken by Homer Simpson.

### Beginner Note: How do you access the Simpsons data?

The `simpsons_script_lines` dataset comes from the `mccoursepack` package you loaded earlier. You can use it just like any other data frame in R. The code below filters for lines spoken by Homer Simpson and randomly selects 100 of them for your sample.

```{r}
#| label: draw-digital-sample
# This code draws a random sample of 100 lines for you

homer_sample <- simpsons_script_lines %>%
  filter(character_name == "Homer Simpson", !is.na(spoken_words)) %>%
  slice_sample(n = 100)

# Let's look at a few lines from our sample
print(homer_sample)
```

### Step 2.2: "Tally" Your Digital Data

Instead of counting dice types, let's count the most common words in our sample of Homer's lines. The code below does this for you. It splits the lines into individual words, removes common "stop words" (like "the", "a", "is"), and counts the rest.

```{r}
#| label: tally-digital-data
# This code tallies the most frequent words in your sample

word_counts <- homer_sample %>%
  unnest_tokens(word, spoken_words) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE)

# Let's see the Top 10 words
print(word_counts)
```

### Step 2.3: Visualize Your Word Sample

Just like with the dice, let's make a bar chart to see the results. This code will plot the top 15 words from your unique sample.

### Beginner Note: Another ggplot2 Example

This code uses `ggplot2` to make a bar chart of the 15 most common words in your sample of Homer Simpson's lines. The bars show how often each word appears. The `coord_flip()` function turns the chart sideways so the words are easier to read.

```{r}
#| label: view-word-plot
# This code creates a bar chart of the most common words in your sample

word_counts %>%
  top_n(15) %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(x = word, y = n)) +
  geom_col(fill = "#FFD90F") + # Springfield yellow!
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Most Common Words in My Homer Simpson Sample",
    x = "Word",
    y = "Frequency in Sample"
  )
```

## 5. Final Reflection

Think about the two parts of this lab. In Part 1, you used a sample of dice to make an inference about the whole bag. In Part 2, you used a sample of words to make an inference about Homer's character.

-   The goal in both cases is **generalizability**: the extent to which we can apply findings from our sample to the entire, unobserved population.
-   How was analyzing the Simpsons word sample similar to analyzing the dice sample? What is the main goal we are trying to achieve in both cases?