---
title: "Week 2: The Publishing Engine"
subtitle: "Building Your Reproducible Workflow"
author: "[Student Name]"
date: last-modified
format:
  html:
    embed-resources: true
    toc: true
    code-fold: false
    code-tools: true
---

## 📖 This Week's Reading

**Textbook**: Chapter 2 - The Publishing Engine

**Key Concepts**:
- The replication crisis and why reproducibility matters
- Code as recipe: literate programming
- The five levels of reproducibility
- R, RStudio, Quarto, and Git as the publishing engine

## 🎯 Learning Objectives

By the end of this week, you will:

- Install and configure R, RStudio, Quarto, and Git
- Understand why code-based workflows prevent replication crises
- Create your first reproducible Quarto document
- Execute basic R commands and document your process

## 🛠️ Setup: Installing the Publishing Engine

### Step 1: Install Required Software

If you haven't already, install:

1. **R** (4.3.0 or higher): <https://cran.r-project.org/>
2. **RStudio Desktop** (free): <https://posit.co/download/rstudio-desktop/>
3. **Quarto** (1.3 or higher): <https://quarto.org/docs/get-started/>
4. **Git**: <https://git-scm.com/downloads>

### Step 2: Install Course Package

```{r}
#| eval: false

# Install remotes package if you don't have it
install.packages("remotes")

# Install your course package (replace with your actual repo)
remotes::install_github("yourprof/coursepackname")

# Load and run setup
library(coursepackname)
coursepack_setup()
```

### Step 3: Verify Installation

```{r}
#| eval: false

# Run self-test
coursepack_self_test()
```

This will check:
- ✅ R version (4.3.0+)
- ✅ RStudio version
- ✅ Quarto installation
- ✅ Git configuration
- ✅ Required R packages

## 📝 Assignment: Your First Reproducible Document

### Part 1: Understanding the Replication Crisis

Answer the following in 2-3 paragraphs:

1. According to Chapter 2, what percentage of psychology studies successfully replicated in the Open Science Collaboration's 2015 study?

[Your answer here]

2. Why can't point-and-click workflows (e.g., Excel → SPSS → Word) guarantee reproducibility?

[Your answer here]

3. Explain the "recipe metaphor" for literate programming. What are the ingredients, instructions, and final dish in a Quarto document?

[Your answer here]

### Part 2: Hands-On R Basics

Execute the following code chunks and observe the output. Then answer the questions.

#### Load Packages

```{r}
#| label: load-packages
#| message: false

library(tidyverse)  # Data wrangling and visualization
```

#### Create Sample Data

Let's create a simple dataset about hypothetical Twitch streamers:

```{r}
#| label: create-data

# Create a data frame
streamers <- tibble(
  username = c("GamerGirl", "ProSniper", "ChillVibes", "SpeedRunner", "CozyQueen"),
  game = c("Valorant", "CS:GO", "Stardew Valley", "Mario", "Animal Crossing"),
  avg_viewers = c(1200, 3500, 800, 2100, 650),
  uses_webcam = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  chat_toxicity = c(6, 8, 2, 5, 1)  # Scale 1-10
)

# Display the data
streamers
```

**Question 1**: How many rows and columns does this dataset have?

[Your answer here]

**Question 2**: Which streamer has the highest average viewership?

[Your answer here]

#### Calculate Descriptive Statistics

```{r}
#| label: descriptive-stats

# Mean viewership
mean(streamers$avg_viewers)

# Median viewership
median(streamers$avg_viewers)

# Summary statistics
summary(streamers$avg_viewers)
```

**Question 3**: Why is the mean (1650) higher than the median (1200)? What does this tell you about the distribution?

[Your answer here]

#### Visualize the Data

```{r}
#| label: plot-viewers
#| fig-cap: "Average viewers by streamer"
#| fig-width: 8
#| fig-height: 5

ggplot(streamers, aes(x = username, y = avg_viewers, fill = game)) +
  geom_col() +
  labs(
    title = "Average Viewership by Streamer",
    x = "Streamer Username",
    y = "Average Viewers",
    fill = "Game Played"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Question 4**: Does there seem to be a relationship between game type and viewership? (Note: This sample is too small for real conclusions, but practice making observations.)

[Your answer here]

#### Test a Hypothesis

Let's test: "Streamers who use webcams have higher viewership."

```{r}
#| label: webcam-analysis

# Compare means
streamers %>%
  group_by(uses_webcam) %>%
  summarize(
    mean_viewers = mean(avg_viewers),
    n = n()
  )
```

**Question 5**: Based on this tiny sample, do webcam users have higher average viewership? What would you need to do to test this properly? (Hint: Think about Chapter 2's discussion of replication.)

[Your answer here]

### Part 3: The One-Click Philosophy

Now, imagine your instructor says: "Actually, we want to add 5 more streamers to the dataset."

In a traditional workflow (Excel + SPSS + Word), you'd have to:
1. Re-import data
2. Re-click through menus
3. Re-copy-paste tables and figures
4. Re-format everything

In this Quarto workflow, you'd:
1. Update the `streamers` tibble in the code chunk
2. Click "Render"
3. Done. Everything updates automatically.

**Reflection** (100-150 words): How does this "one-click" philosophy change the way you think about research? Why does it matter that others can verify your work?

[Your reflection here]

## 🧪 Experimentation Challenge (Optional Bonus)

Modify the `streamers` dataset to add yourself as a hypothetical streamer:

- Choose a username
- Pick a game you'd stream
- Estimate your avg_viewers (be realistic or aspirational)
- Decide if you'd use a webcam
- Rate your expected chat_toxicity (1-10 scale)

Then re-render the document and observe how the statistics and plot update automatically.

**What changed?**

[Your observations here]

## 💭 Discussion Questions

1. Chapter 2 distinguishes five levels of reproducibility (Publication Only → Literate Executable Documents). At what level does most undergraduate coursework currently operate? Why?

2. The textbook argues that "computational literacy is media literacy." Do you agree? How does learning R help you understand algorithmic media?

3. If someone can't reproduce your research, does that necessarily mean it's wrong? What are the implications of non-reproducible findings for journalism, policymaking, or academic knowledge?

## ✅ Submission Checklist

- [ ] All code chunks execute without errors (click "Render" to test)
- [ ] All questions answered with complete sentences
- [ ] Figures display correctly in the rendered HTML
- [ ] File naming: `Week02_[YourLastName].qmd` and `Week02_[YourLastName].html`
- [ ] Submitted both source (.qmd) and output (.html)

## 📚 Resources

- **Textbook**: Chapter 2 (full_textbook.md, lines 240-694)
- **R for Data Science**: <https://r4ds.hadley.nz/> (free online book)
- **Quarto Documentation**: <https://quarto.org/docs/guide/>
- **Cheat Sheets**: RStudio → Help → Cheat Sheets

---

**Estimated Time**: 3-4 hours (including software installation)  
**Due**: [Instructor specifies]  
**Points**: [Instructor specifies]
