---
title: "MC 451 R Workbook"
subtitle: "A guided tour of R — from wrangling to publishing"
author: "MC 451 · Research Methods in Mass Media · Spring 2026"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-title: "Sections"
    number-sections: true
    smooth-scroll: true
    code-fold: false
    code-tools: true
    code-copy: true
    highlight-style: github
    fig-align: center
    cap-location: bottom
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

<!-- ═══════════════════════════════════════════════════════════════
     HOW TO USE THIS WORKBOOK

     1. Open mc451-r-workbook.Rproj in RStudio first.
     2. Work top to bottom. Watch each video before running that section.
     3. Run code chunks with Ctrl+Shift+Enter (Windows) or Cmd+Shift+Enter (Mac).
     4. "Try It" boxes are small guided challenges — do them before moving on.
     5. Nothing here is graded. Experiment freely.
     ═══════════════════════════════════════════════════════════════ -->

::: {.callout-important}
## Before You Begin

Complete the README setup steps first:

1. Open this file via `mc451-r-workbook.Rproj` (not by double-clicking the .qmd)
2. Run the package install chunk below **once**
3. Watch Video 1, then start Section 1
:::

---

# Setup {.unnumbered}

::: {.callout-tip}
## 📹 Video 1 — Getting Started
*Watch this before running anything. (~5 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-1)

*Covers: Opening an RStudio Project, running code chunks, how this workbook connects to your assignments and final portfolio.*
:::

## Install Packages (Run Once)

```{r}
#| label: install-packages
#| eval: false

# Run this chunk ONE TIME. After packages are installed, you never need to run it again.
install.packages(c(
  "tidyverse",   # dplyr, ggplot2, readr, and more
  "janitor",     # clean_names() and other data cleaning tools
  "kableExtra",  # publication-quality tables
  "scales",      # axis formatting helpers
  "effsize"      # effect size calculations (Cohen's d)
))
```

## Load Packages and Data

Run this chunk every time you open the workbook. It loads everything you need.

```{r}
#| label: setup

library(tidyverse)
library(janitor)
library(kableExtra)
library(scales)
library(effsize)
library(coursepackR)

# Load the course dataset
data(unified_music)
```

## First Look at the Data

```{r}
#| label: first-look

# How many rows and columns?
nrow(unified_music)
ncol(unified_music)

# What are the column names?
names(unified_music)

# Quick snapshot of structure and types
glimpse(unified_music)
```

::: {.callout-note}
## What You Should See

- **1,792 rows** — one per song
- **36 columns** — Billboard chart data, Spotify audio features, Genius metadata
- Mixed column name styles (`Song` vs. `danceability`) — we fix this in Section 1
- Several `<chr>`, `<int>`, and `<dbl>` column types
:::

## How This Workbook Connects to Your Work

| This workbook... | Prepares you for... |
|---|---|
| Section 1: Wrangling | R Assignment 01 |
| Section 2: Describing | R Assignment 02 |
| Section 3: Testing | R Assignment 03 |
| Section 4: Interpreting | R Assignment 03 + Final Portfolio |
| Section 5: Portfolio | Final Portfolio (writing + rendering) |
| Section 6: Publishing | Final Portfolio (GitHub Pages) |

The dataset used here (`unified_music`) is the clean, packaged version. Your assignments use the raw CSV (`music_data_raw.csv`) — same data, same skills.

---

# Chapter 17 Recap — Wrangling the Chaos {#sec-wrangling}

::: {.callout-tip}
## 📹 Video 2 — Wrangling the Chaos
*Watch before running this section. (~8 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-2)

*Covers: The pipe operator, inspecting data, the five core dplyr verbs, saving clean data.*
:::

We covered Chapter 17 in class today. This section reinforces those concepts with hands-on code you can run and modify.

## The Pipe Operator `|>`

The pipe passes the result of one operation into the next. Instead of nesting functions:

```{r}
#| label: pipe-demo

# Without the pipe — hard to read
arrange(filter(unified_music, playlist_genre == "pop"), desc(Weeks_on_Board))

# With the pipe — reads left to right, top to bottom
unified_music |>
  filter(playlist_genre == "pop") |>
  arrange(desc(Weeks_on_Board))
```

::: {.callout-note}
## The Shortcut

Insert `|>` with **Ctrl+Shift+M** (Windows) or **Cmd+Shift+M** (Mac).

You may also see `%>%` in older code — it works the same way. `|>` is the modern native R version.
:::

## Cleaning Column Names

The `unified_music` dataset has inconsistent naming (`Song` vs. `danceability`). `clean_names()` standardizes everything to `snake_case`.

```{r}
#| label: clean-names

# Before: mixed case column names
names(unified_music)

# Apply clean_names() and save as a new object
music <- unified_music |>
  clean_names()

# After: all lowercase, underscores instead of spaces/capitals
names(music)
```

::: {.callout-important}
## Use `music` from Here On

After running the chunk above, use `music` (not `unified_music`) for the rest of the workbook. `music` has clean column names that are much easier to type.
:::

## Inspecting Data

```{r}
#| label: inspect

# Overall structure
glimpse(music)

# First 6 rows
head(music)

# Last 6 rows
tail(music)

# Summary statistics for every column
summary(music)

# Count values in a categorical column
music |> count(playlist_genre)

# Count and sort
music |> count(playlist_genre, sort = TRUE)
```

## Selecting Columns

```{r}
#| label: select

# Keep only specific columns
music |>
  select(song, artist, playlist_genre, danceability, energy, valence)

# Keep a range of columns
music |>
  select(song:artist, danceability:valence)

# Drop a column with -
music |>
  select(-track_id, -playlist_id, -track_album_id)
```

## Filtering Rows

```{r}
#| label: filter

# Songs from a single genre
music |>
  filter(playlist_genre == "pop")

# Songs that charted at #1
music |>
  filter(max_rank == 1)

# Songs with high danceability (top quarter, roughly)
music |>
  filter(danceability > 0.75)

# Multiple conditions: pop songs with high valence
music |>
  filter(playlist_genre == "pop", valence > 0.7)

# Songs NOT in pop
music |>
  filter(playlist_genre != "pop")
```

::: {.callout-note}
## Try It

Find all songs that stayed on the Billboard chart for more than 30 weeks. How many are there? Which genre has the most?

```{r}
#| label: try-filter

# Your code here
music |>
  filter(weeks_on_board > 30) |>
  count(playlist_genre, sort = TRUE)
```
:::

## Creating and Modifying Columns

```{r}
#| label: mutate

# Create a new column: song duration in minutes
music <- music |>
  mutate(duration_min = duration_ms / 60000)

# Recode a numeric column to meaningful labels
music <- music |>
  mutate(mode_label = case_when(
    mode == 1 ~ "major",
    mode == 0 ~ "minor",
    TRUE      ~ NA_character_
  ))

# Create a decade column from year
music <- music |>
  mutate(decade = floor(year / 10) * 10)

# Verify your new columns
music |>
  select(song, year, decade, mode, mode_label, duration_ms, duration_min) |>
  head(10)
```

::: {.callout-note}
## What `case_when()` Does

`case_when()` is a flexible way to recode values. Each line is a condition `~` result pair. The last line (`TRUE ~ ...`) is the catch-all for anything that didn't match a previous condition — like an `else` statement.
:::

## Handling Missing Values

```{r}
#| label: missing

# Find columns with missing values
music |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

# Remove rows where a specific column is missing
music_no_missing_views <- music |>
  filter(!is.na(views))

nrow(music)              # before
nrow(music_no_missing_views)  # after

# Or: drop rows missing ANY column
music |>
  drop_na()
```

## Summarizing Data

```{r}
#| label: summarise

# Overall summary statistics
music |>
  summarise(
    mean_danceability = mean(danceability, na.rm = TRUE),
    sd_danceability   = sd(danceability, na.rm = TRUE),
    min_danceability  = min(danceability, na.rm = TRUE),
    max_danceability  = max(danceability, na.rm = TRUE)
  )

# Summary statistics by group
music |>
  group_by(playlist_genre) |>
  summarise(
    n             = n(),
    mean_valence  = mean(valence, na.rm = TRUE),
    mean_energy   = mean(energy, na.rm = TRUE),
    mean_weeks    = mean(weeks_on_board, na.rm = TRUE)
  ) |>
  arrange(desc(mean_valence))
```

## Sorting and Slicing

```{r}
#| label: arrange-slice

# Sort by chart performance (best rank first)
music |>
  arrange(max_rank) |>
  select(song, artist, max_rank, weeks_on_board) |>
  head(10)

# Sort descending
music |>
  arrange(desc(weeks_on_board)) |>
  select(song, artist, weeks_on_board) |>
  head(10)

# Grab top 5 within each genre
music |>
  group_by(playlist_genre) |>
  slice_max(weeks_on_board, n = 5) |>
  select(song, artist, playlist_genre, weeks_on_board)
```

## Saving Your Work

```{r}
#| label: save

# Save as RDS (preserves R data types — use for passing between R scripts)
saveRDS(music, "music_workbook_clean.RDS")

# Save as CSV (readable in Excel — use for sharing or archiving)
write_csv(music, "music_workbook_clean.csv")

# Read back in
music_reload <- readRDS("music_workbook_clean.RDS")
```

::: {.callout-note}
## RDS vs CSV

Use **RDS** when the file will be read back into R — it preserves factors, date types, and other R-specific structures exactly. Use **CSV** when you need to share the file with someone using Excel or another program.
:::

---

# Chapter 18 — Seeing Patterns with ggplot2 {#sec-describing}

::: {.callout-tip}
## 📹 Video 3 — Seeing Patterns
*Watch before running this section. (~10 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-3)

*Covers: The grammar of graphics, building bar charts and histograms, customizing with labels and themes, saving plots.*
:::

## The Grammar of Graphics

Every ggplot is built from three required pieces:

```{r}
#| label: ggplot-anatomy

# Template:
# ggplot(data = YOUR_DATA, aes(x = VAR1, y = VAR2)) +
#   geom_SOMETHING()

# Minimal working example
ggplot(data = music, aes(x = playlist_genre)) +
  geom_bar()
```

The `+` adds layers — aesthetics, geometries, labels, themes. Think of it like building a sandwich: each `+` adds another ingredient.

## Bar Chart — Categorical Variable

```{r}
#| label: bar-chart

# How many songs per genre?
music |>
  count(playlist_genre) |>
  ggplot(aes(x = playlist_genre, y = n, fill = playlist_genre)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Songs per Genre in the Unified Music Dataset",
    subtitle = "Source: Billboard Hot 100 merged with Spotify & Genius data",
    x = "Genre",
    y = "Number of Songs"
  ) +
  theme_minimal()
```

```{r}
#| label: bar-chart-ordered

# Same chart, ordered by count (more informative)
music |>
  count(playlist_genre) |>
  mutate(playlist_genre = fct_reorder(playlist_genre, n)) |>
  ggplot(aes(x = playlist_genre, y = n, fill = playlist_genre)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +                        # horizontal bars are easier to read
  labs(
    title = "Songs per Genre",
    x = NULL,
    y = "Number of Songs"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Try It

Make a bar chart showing the number of songs per **decade**. Which decade has the most songs in the dataset?

```{r}
#| label: try-bar

# Your code here
music |>
  count(decade) |>
  ggplot(aes(x = factor(decade), y = n)) +
  geom_col() +
  labs(title = "Songs per Decade", x = "Decade", y = "Count") +
  theme_minimal()
```
:::

## Histogram — Continuous Variable

```{r}
#| label: histogram

# Distribution of danceability scores
ggplot(music, aes(x = danceability)) +
  geom_histogram(bins = 30, fill = "#b71c1c", color = "white") +
  labs(
    title = "Distribution of Song Danceability",
    x = "Danceability (0–1)",
    y = "Number of Songs"
  ) +
  theme_minimal()
```

```{r}
#| label: histogram-valence

# Distribution of valence (musical positivity)
ggplot(music, aes(x = valence)) +
  geom_histogram(bins = 30, fill = "#1565c0", color = "white") +
  geom_vline(xintercept = mean(music$valence), linetype = "dashed", color = "black") +
  labs(
    title = "Distribution of Song Valence",
    subtitle = "Dashed line = dataset mean",
    x = "Valence (0 = sad/dark, 1 = happy/bright)",
    y = "Number of Songs"
  ) +
  theme_minimal()
```

## Box Plot — Comparing Groups

```{r}
#| label: boxplot

# Valence by genre — do some genres skew happier?
ggplot(music, aes(x = playlist_genre, y = valence, fill = playlist_genre)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8) +
  labs(
    title = "Song Valence by Genre",
    x = "Genre",
    y = "Valence (musical positivity)"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Reading a Box Plot

- **Center line** = median
- **Box edges** = 25th and 75th percentile (the middle 50% of data)
- **Whiskers** = extend to 1.5× the interquartile range
- **Dots beyond whiskers** = outliers
:::

## Scatter Plot — Two Continuous Variables

```{r}
#| label: scatter

# Do energetic songs tend to be danceable?
ggplot(music, aes(x = energy, y = danceability)) +
  geom_point(alpha = 0.3, color = "#b71c1c", size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  labs(
    title = "Energy vs. Danceability",
    subtitle = "Each point = one song | Line = linear trend",
    x = "Energy (0–1)",
    y = "Danceability (0–1)"
  ) +
  theme_minimal()
```

```{r}
#| label: scatter-color

# Same plot, colored by genre
ggplot(music, aes(x = energy, y = danceability, color = playlist_genre)) +
  geom_point(alpha = 0.4, size = 1.5) +
  labs(
    title = "Energy vs. Danceability by Genre",
    x = "Energy",
    y = "Danceability",
    color = "Genre"
  ) +
  theme_minimal()
```

## Faceting — Small Multiples

```{r}
#| label: facet

# One histogram per genre, arranged in a grid
ggplot(music, aes(x = valence)) +
  geom_histogram(bins = 20, fill = "#b71c1c", color = "white") +
  facet_wrap(~ playlist_genre, ncol = 3) +
  labs(
    title = "Valence Distribution by Genre",
    x = "Valence",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: facet-decade

# Danceability trends by decade
music |>
  filter(!is.na(decade)) |>
  ggplot(aes(x = danceability)) +
  geom_histogram(bins = 20, fill = "#1565c0", color = "white") +
  facet_wrap(~ decade, ncol = 4) +
  labs(
    title = "Danceability by Decade",
    x = "Danceability",
    y = "Count"
  ) +
  theme_minimal(base_size = 10)
```

## Customizing Plots

```{r}
#| label: custom-plot

# A polished, publication-ready bar chart
music |>
  count(playlist_genre) |>
  mutate(
    pct = n / sum(n),
    playlist_genre = fct_reorder(playlist_genre, pct)
  ) |>
  ggplot(aes(x = playlist_genre, y = pct, fill = playlist_genre)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    hjust = -0.15,
    size = 3.5
  ) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 0.35)
  ) +
  scale_fill_manual(values = c(
    "#b71c1c", "#e53935", "#ef9a9a",
    "#1565c0", "#1976d2", "#64b5f6"
  )) +
  coord_flip() +
  labs(
    title = "Genre Distribution in the Unified Music Dataset",
    subtitle = "Billboard Hot 100 songs merged with Spotify & Genius data | N = 1,792",
    x = NULL,
    y = "Share of Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "#666666", size = 9),
    panel.grid.major.y = element_blank()
  )
```

## Saving Plots

```{r}
#| label: save-plot

# Save the most recent plot
ggsave(
  filename = "genre_distribution.png",
  width    = 7,
  height   = 4.5,
  dpi      = 300
)

# Or save a specific named plot object
my_plot <- ggplot(music, aes(x = energy, y = valence)) +
  geom_point(alpha = 0.3)

ggsave("energy_valence.png", plot = my_plot, width = 6, height = 4, dpi = 300)
```

::: {.callout-note}
## Try It

Create a scatter plot showing the relationship between `tempo` and `energy`. Add a linear trend line. Does tempo predict energy?

```{r}
#| label: try-scatter

# Your code here
ggplot(music, aes(x = tempo, y = energy)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Tempo vs. Energy", x = "Tempo (BPM)", y = "Energy") +
  theme_minimal()
```
:::

---

# Chapter 19 — The Surprise Detector: Statistical Tests {#sec-testing}

::: {.callout-tip}
## 📹 Video 4 — The Surprise Detector
*Watch before running this section. (~10 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-4)

*Covers: What a p-value actually means, the two-sample t-test, chi-square test, reading R output, APA reporting.*
:::

Statistical tests answer one question: **Could this pattern be due to chance?**

If the answer is "probably not" (p < .05), we call the result statistically significant.

## The Two-Sample T-Test

Use a t-test when you want to compare the **means** of a **continuous variable** across **two groups**.

**Research question:** Do pop songs have significantly higher valence (musical positivity) than rock songs?

```{r}
#| label: ttest-setup

# Step 1: Filter to just the two groups
pop_rock <- music |>
  filter(playlist_genre %in% c("pop", "rock"))

# Step 2: Check the group sizes
pop_rock |> count(playlist_genre)

# Step 3: Check group means visually
pop_rock |>
  group_by(playlist_genre) |>
  summarise(
    n    = n(),
    mean = mean(valence),
    sd   = sd(valence)
  )
```

```{r}
#| label: ttest-run

# Step 4: Run the t-test
t_result <- t.test(valence ~ playlist_genre, data = pop_rock)

# Step 5: Read the output
t_result
```

::: {.callout-note}
## Reading t-test Output

```
t = 5.23, df = 498, p-value < 0.001
95% CI: [0.08, 0.17]
```

- **t** — the test statistic (how many standard errors apart the means are)
- **df** — degrees of freedom (related to sample size)
- **p-value** — the probability of seeing a difference this large if there were truly no difference
- **95% CI** — the range we're 95% confident contains the true difference in means

**APA format:** *t*(498) = 5.23, *p* < .001
:::

```{r}
#| label: ttest-extract

# Extract values for inline reporting
t_stat <- round(t_result$statistic, 2)
t_df   <- round(t_result$parameter, 0)
t_p    <- round(t_result$p.value, 3)
t_ci   <- round(t_result$conf.int, 2)

# Print a formatted result
cat("t(", t_df, ") =", t_stat, ", p =", t_p)
cat("\n95% CI: [", t_ci[1], ",", t_ci[2], "]")
```

```{r}
#| label: ttest-viz

# Visualize the comparison
ggplot(pop_rock, aes(x = playlist_genre, y = valence, fill = playlist_genre)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.8, width = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "white") +
  scale_fill_manual(values = c("#b71c1c", "#1565c0")) +
  labs(
    title = "Valence: Pop vs. Rock",
    subtitle = "Diamond = group mean",
    x = "Genre",
    y = "Valence (musical positivity)"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Try It

Run a t-test comparing the **danceability** of `r&b` songs vs. `rap` songs. Are they significantly different?

```{r}
#| label: try-ttest

# Your code here
rb_rap <- music |> filter(playlist_genre %in% c("r&b", "rap"))
t.test(danceability ~ playlist_genre, data = rb_rap)
```
:::

## The Chi-Square Test

Use a chi-square test when both variables are **categorical** (e.g., genre × mode). This is the test you will use in your **final portfolio**.

**Research question:** Is there a relationship between genre and mode (major vs. minor)?

```{r}
#| label: chisq-setup

# Step 1: Make sure the variables are factors with meaningful labels
music_chi <- music |>
  mutate(
    mode_label    = factor(mode_label, levels = c("major", "minor")),
    playlist_genre = factor(playlist_genre)
  ) |>
  filter(!is.na(mode_label))

# Step 2: Cross-tabulation
cross_tab <- table(music_chi$playlist_genre, music_chi$mode_label)
cross_tab

# Step 3: Add proportions
prop.table(cross_tab, margin = 1) |> round(3)  # row proportions
```

```{r}
#| label: chisq-run

# Step 4: Run the chi-square test
chi_result <- chisq.test(cross_tab)
chi_result

# Step 5: Check expected cell counts (should all be ≥ 5)
chi_result$expected
min(chi_result$expected)
```

```{r}
#| label: chisq-extract

# Extract values
chi_x2 <- round(chi_result$statistic, 2)
chi_df <- chi_result$parameter
chi_p  <- round(chi_result$p.value, 3)

cat("χ²(", chi_df, ") =", chi_x2, ", p =", chi_p)
```

::: {.callout-note}
## Reading Chi-Square Output

```
X-squared = 14.3, df = 5, p-value = 0.014
```

- **X-squared** — the chi-square statistic
- **df** — degrees of freedom (number of rows − 1) × (number of columns − 1)
- **p-value** — probability of this association occurring by chance alone

**APA format:** χ²(5) = 14.30, *p* = .014

**Assumption check:** All expected cell counts must be ≥ 5. Check `chi_result$expected`.
:::

```{r}
#| label: chisq-viz

# Stacked bar chart: proportion of major/minor by genre
music_chi |>
  count(playlist_genre, mode_label) |>
  group_by(playlist_genre) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = playlist_genre, y = pct, fill = mode_label)) +
  geom_col(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("#1565c0", "#b71c1c")) +
  labs(
    title = "Major vs. Minor Key by Genre",
    x     = "Genre",
    y     = "Proportion",
    fill  = "Mode"
  ) +
  theme_minimal()
```

## Pearson Correlation

Use correlation when both variables are **continuous** and you want to measure how strongly they move together.

**Research question:** Is there a relationship between energy and loudness?

```{r}
#| label: correlation

# Run the correlation test
cor_result <- cor.test(music$energy, music$loudness)
cor_result

# Extract values
r_val <- round(cor_result$estimate, 2)
r_p   <- round(cor_result$p.value, 3)
r_ci  <- round(cor_result$conf.int, 2)

cat("r =", r_val, ", p =", r_p)
cat("\n95% CI: [", r_ci[1], ",", r_ci[2], "]")
```

::: {.callout-note}
## Interpreting Pearson's r

| r value | Interpretation |
|---------|---------------|
| .00–.09 | Negligible |
| .10–.29 | Small |
| .30–.49 | Moderate |
| .50–.69 | Large |
| .70–1.0 | Very large |

**Direction:** Positive r = as one variable increases, so does the other. Negative r = as one increases, the other decreases.

**APA format:** *r*(1790) = .78, *p* < .001
:::

```{r}
#| label: correlation-viz

ggplot(music, aes(x = energy, y = loudness)) +
  geom_point(alpha = 0.2, color = "#b71c1c", size = 1.2) +
  geom_smooth(method = "lm", color = "black") +
  labs(
    title    = paste0("Energy vs. Loudness (r = ", r_val, ")"),
    subtitle = paste0("p = ", r_p, " | N = ", nrow(music)),
    x = "Energy",
    y = "Loudness (dB)"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Try It

Test the correlation between `danceability` and `valence`. Is the relationship significant? How would you describe the strength?

```{r}
#| label: try-correlation

# Your code here
cor.test(music$danceability, music$valence)
```
:::

---

# Chapter 20 — Interpreting the Call: Effect Sizes {#sec-interpreting}

::: {.callout-tip}
## 📹 Video 5 — Interpreting the Call
*Watch before running this section. (~8 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-5)

*Covers: Why p-values are not enough, Cohen's d, Cramér's V, writing honest interpretations.*
:::

A p-value tells you **whether** an effect exists. An effect size tells you **how big** it is. Both are required for a complete interpretation.

## Cohen's d — Effect Size for T-Tests

```{r}
#| label: cohens-d

# Cohen's d for the pop vs. rock valence comparison
d_result <- cohen.d(
  valence ~ playlist_genre,
  data = pop_rock
)

d_result

# Extract the d value
d_val <- round(d_result$estimate, 2)
cat("Cohen's d =", d_val)
```

::: {.callout-note}
## Cohen's d Conventions

| d value | Label |
|---------|-------|
| 0.2 | Small |
| 0.5 | Medium |
| 0.8 | Large |

A significant p-value with d = 0.1 means the difference is real but tiny. A significant p-value with d = 0.8 means the difference is both real and large.
:::

## Cramér's V — Effect Size for Chi-Square

```{r}
#| label: cramers-v

# Calculate Cramér's V manually from the chi-square result
n       <- sum(cross_tab)
k       <- min(nrow(cross_tab), ncol(cross_tab))
cramers_v <- sqrt(chi_result$statistic / (n * (k - 1)))
cramers_v <- round(cramers_v, 3)

cat("Cramér's V =", cramers_v)
```

::: {.callout-note}
## Cramér's V Conventions

| V value | Label |
|---------|-------|
| .10 | Small |
| .30 | Medium |
| .50 | Large |

Cramér's V ranges from 0 (no association) to 1 (perfect association).
:::

## Statistical vs. Practical Significance

```{r}
#| label: stat-vs-practical

# With a large sample (N = 1,792), even tiny differences can be significant
# Let's see what the actual mean difference is

pop_rock |>
  group_by(playlist_genre) |>
  summarise(mean_valence = mean(valence))

# The t-test may be significant, but is the difference meaningful?
# This is where effect sizes come in.
```

::: {.callout-note}
## A Key Distinction

**Statistical significance** (p < .05): The pattern is unlikely to be due to chance.

**Practical significance** (effect size): The pattern is large enough to matter in the real world.

Large samples (like ours with N = 1,792) detect even trivial differences as "significant." Always report and interpret the effect size alongside the p-value.
:::

## Writing Results in APA Format

Here are templates for each test type. Copy and adapt for your own work.

**T-test result:**
> A two-sample t-test revealed that pop songs (*M* = 0.61, *SD* = 0.19) had significantly higher valence than rock songs (*M* = 0.47, *SD* = 0.22), *t*(498) = 5.23, *p* < .001, *d* = 0.46. This represents a medium effect, suggesting that genre differences in emotional tone are practically meaningful, not merely statistically detectable.

**Correlation result:**
> Energy and loudness were strongly and positively correlated, *r*(1790) = .78, *p* < .001, 95% CI [.75, .80]. Songs rated higher in energy by Spotify's algorithm were consistently recorded at higher loudness levels, which aligns with the acoustic logic underlying both metrics.

**Chi-square result:**
> A chi-square test of independence revealed a significant association between genre and mode, χ²(5) = 14.30, *p* = .014, *V* = .09. However, the effect size is small, indicating that while the association is real, genre accounts for very little of the variation in whether songs are written in a major or minor key.

```{r}
#| label: write-results

# Useful inline values for writing results
pop_mean  <- music |> filter(playlist_genre == "pop")  |> pull(valence) |> mean() |> round(2)
rock_mean <- music |> filter(playlist_genre == "rock") |> pull(valence) |> mean() |> round(2)
pop_sd    <- music |> filter(playlist_genre == "pop")  |> pull(valence) |> sd()   |> round(2)
rock_sd   <- music |> filter(playlist_genre == "rock") |> pull(valence) |> sd()   |> round(2)

cat("Pop: M =", pop_mean, ", SD =", pop_sd)
cat("\nRock: M =", rock_mean, ", SD =", rock_sd)
```

::: {.callout-note}
## Try It

For the r&b vs. rap danceability t-test you ran earlier, calculate Cohen's d and write a one-sentence APA-format interpretation that includes both the t-test result and the effect size.

```{r}
#| label: try-effect-size

# Your code here
rb_rap <- music |> filter(playlist_genre %in% c("r&b", "rap"))
cohen.d(danceability ~ playlist_genre, data = rb_rap)
```
:::

---

# Chapter 21 — Building Your Portfolio Document {#sec-portfolio}

::: {.callout-tip}
## 📹 Video 6 — Building Your Portfolio Document
*Watch before running this section. (~10 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-6)

*Covers: Quarto document anatomy, chunk options, inline R, kableExtra tables, rendering to HTML and PDF.*
:::

Your final portfolio is a Quarto Book — the same format as this workbook. This section teaches you the Quarto skills you need to write it.

## Quarto Document Anatomy

A Quarto document has three parts:

```
---
title: "My Research Report"     ← YAML header (controls format/settings)
author: "Your Name"
format: html
---

Some text here.                 ← Markdown text (prose)

```{r}                          ← Code chunk (R code)
1 + 1
```
```

## YAML Header Options

```{r}
#| label: yaml-demo
#| eval: false

# This is a typical YAML header for your final portfolio sections
# (already set up in mc451-final-project/_quarto.yml)

# Key options:
# toc: true           → table of contents
# number-sections: true → section numbering
# code-fold: true     → collapsible code (readers can choose to see it)
# echo: false         → hide code by default (show only output)
# warning: false      → suppress warning messages in output
```

## Code Chunk Options

Control how each chunk behaves using `#|` comments at the top:

```{r}
#| label: chunk-options-demo
#| echo: true          # show the code in the output? (true/false)
#| eval: true          # actually run the code? (true/false)
#| warning: false      # suppress warnings?
#| message: false      # suppress messages?
#| fig-width: 7        # figure width in inches
#| fig-height: 4       # figure height in inches
#| fig-cap: "This caption appears below the figure in the rendered document."

ggplot(music, aes(x = danceability)) +
  geom_histogram(bins = 30, fill = "#b71c1c") +
  theme_minimal()
```

## Inline R Code

Instead of typing numbers manually, use inline R to pull values directly from your analysis:

In the .qmd file, write:

```
The dataset contains `r nrow(music)` songs across `r n_distinct(music$playlist_genre)` genres.
```

This renders as:

> The dataset contains 1792 songs across 6 genres.

If your data changes, the numbers update automatically. **Never type statistical results by hand** — always use inline R.

```{r}
#| label: inline-examples

# Values you might use inline in your final portfolio
n_songs    <- nrow(music)
n_genres   <- n_distinct(music$playlist_genre)
mean_dance <- round(mean(music$danceability), 2)

# In your .qmd text you would write:
# "The sample consists of `r n_songs` songs..."
# "Mean danceability was `r mean_dance`..."
```

## Publication Tables with kableExtra

```{r}
#| label: kable-basic

# Basic frequency table
music |>
  count(playlist_genre, name = "n") |>
  mutate(
    Percent = round(n / sum(n) * 100, 1),
    Percent = paste0(Percent, "%")
  ) |>
  rename(Genre = playlist_genre, N = n) |>
  kable(caption = "Table 1. Songs by Genre", align = c("l", "c", "c")) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width        = FALSE
  )
```

```{r}
#| label: kable-styled

# A more polished table with a colored header
music |>
  group_by(playlist_genre) |>
  summarise(
    N             = n(),
    Mean_Valence  = round(mean(valence), 2),
    Mean_Energy   = round(mean(energy), 2),
    Mean_Weeks    = round(mean(weeks_on_board), 1)
  ) |>
  rename(
    Genre = playlist_genre,
    `Mean Valence` = Mean_Valence,
    `Mean Energy`  = Mean_Energy,
    `Mean Weeks on Chart` = Mean_Weeks
  ) |>
  kable(caption = "Table 2. Audio Feature Averages by Genre") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) |>
  row_spec(0, bold = TRUE, background = "#b71c1c", color = "white")
```

## Rendering Your Document

```{r}
#| label: render-instructions
#| eval: false

# In the RStudio TERMINAL tab (not the Console), run:
quarto render mc451-r-workbook.qmd

# For your final portfolio:
quarto render   # renders the entire book

# Check the output in the docs/ folder
# Open docs/mc451-r-workbook.html in your browser
```

::: {.callout-important}
## Render This Workbook Now

Switch to the RStudio **Terminal** tab (bottom pane, next to Console) and run:

```bash
quarto render mc451-r-workbook.qmd
```

Open `docs/mc451-r-workbook.html` in your browser. This is exactly what you will do for your final portfolio.
:::

## Common Rendering Errors

| Error | Fix |
|---|---|
| `object 'X' not found` | You skipped a chunk that created `X`. Run all chunks in order. |
| Package not found | Run `install.packages("PACKAGE")` in the Console |
| `could not find function` | The package is installed but not loaded. Add `library(PACKAGE)` to setup chunk |
| Blank output | Your `eval: false` chunk option is preventing execution. Remove it. |
| `figure margins too large` | Increase `fig-width` or `fig-height` in the chunk options |

---

# Chapter 22 — Going Live: GitHub Pages {#sec-publishing}

::: {.callout-tip}
## 📹 Video 7 — Going Live
*Watch before running this section. (~10 min)*

[▶ Watch on YouTube](https://youtu.be/VIDEO-ID-7)

*Covers: Git basics, setting up a GitHub repository, connecting RStudio to GitHub, enabling GitHub Pages, the render → push cycle.*
:::

Your final portfolio renders into the `docs/` folder. GitHub Pages serves that folder as a live website. This section walks you through the full workflow.

## Git Basics

Git tracks the history of your files. Every time you `commit`, you save a snapshot.

The three commands you need:

```bash
git add .                          # Stage all changed files
git commit -m "Brief description"  # Save a snapshot with a message
git push                           # Upload to GitHub
```

You run these in the RStudio **Terminal** tab, not the Console.

## Step 1 — Initialize Git in Your Project

Do this once, in the Terminal, when you're in your project folder:

```bash
git init
git add .
git commit -m "Initial commit"
```

::: {.callout-note}
## Are You in the Right Folder?

In the Terminal, type `pwd` to see your current directory. It should end with `mc451-final-project`. If not, navigate there first or use **Session → Set Working Directory → To Project Directory** in RStudio.
:::

## Step 2 — Create a GitHub Repository

1. Go to [github.com](https://github.com) and sign in
2. Click the **+** → **New repository**
3. Name it `mc451-portfolio` (or anything you prefer)
4. Set to **Public**
5. Do **not** initialize with a README
6. Click **Create repository**

GitHub will show you a page with setup instructions. You need the URL — it looks like:
`https://github.com/YOUR-USERNAME/mc451-portfolio.git`

## Step 3 — Connect Your Project to GitHub

In the RStudio Terminal:

```bash
git remote add origin https://github.com/YOUR-USERNAME/mc451-portfolio.git
git branch -M main
git push -u origin main
```

Replace `YOUR-USERNAME` and `mc451-portfolio` with your actual GitHub username and repo name.

## Step 4 — Enable GitHub Pages

1. Go to your repository on GitHub
2. Click **Settings** → **Pages** (left sidebar)
3. Under **Source**, choose: **Deploy from a branch**
4. Branch: `main` | Folder: `/docs`
5. Click **Save**

Wait 2–3 minutes, then visit:
`https://YOUR-USERNAME.github.io/mc451-portfolio/`

## Step 5 — The Render → Push Cycle

Every time you update your portfolio, repeat these three steps:

```bash
# Step 1: Render (in Terminal)
quarto render

# Step 2: Stage and commit
git add .
git commit -m "Update: describe what you changed"

# Step 3: Push to GitHub
git push
```

GitHub Pages updates your live site within ~2 minutes of each push.

## Step 6 — Verify Your Site

Before submitting:

1. Open your site URL in a **private/incognito browser window**
2. Confirm the site is public (you can see it without being logged into GitHub)
3. Click through every page — Introduction, Methodology, Results, Discussion, References
4. Confirm all tables and figures load correctly
5. Check the PDF link works

::: {.callout-important}
## Common GitHub Problems

| Problem | Fix |
|---|---|
| `Permission denied (publickey)` | Set up SSH keys, or switch to HTTPS: `git remote set-url origin https://github.com/USERNAME/REPO.git` |
| Pages says "not found" | Wait 5 minutes and refresh. Check that `/docs` folder exists and has `index.html` inside. |
| Site shows old content | You forgot to `quarto render` before pushing. Re-render and push again. |
| "Everything up to date" but site hasn't changed | You haven't committed. Run `git add .` then `git commit` before `git push`. |
| Figures missing on live site | You have absolute file paths in your code. Use relative paths only. |
:::

## Practice Run — Using This Workbook

Before your final portfolio, practice the full workflow with this workbook:

```bash
# 1. Render the workbook
quarto render mc451-r-workbook.qmd

# 2. Initialize git (if not done yet)
git init
git add .
git commit -m "Add MC 451 R Workbook"

# 3. Create a practice repo on GitHub called mc451-workbook-practice
# Then connect and push:
git remote add origin https://github.com/YOUR-USERNAME/mc451-workbook-practice.git
git branch -M main
git push -u origin main

# 4. Enable GitHub Pages from /docs
# 5. Visit your live workbook at https://YOUR-USERNAME.github.io/mc451-workbook-practice/
```

::: {.callout-tip}
## You're Ready

If you can see this workbook live on GitHub Pages, you have every skill you need to publish your final portfolio. The final portfolio workflow is identical — just with your own research content instead.
:::

---

# Quick Reference {.unnumbered}

## Packages Used in This Course

| Package | Purpose | Key Functions |
|---------|---------|---------------|
| `tidyverse` | Data import, wrangling, visualization | `read_csv()`, `filter()`, `mutate()`, `ggplot()` |
| `janitor` | Column name cleaning | `clean_names()`, `get_dupes()` |
| `kableExtra` | Publication tables | `kable()`, `kable_styling()`, `row_spec()` |
| `scales` | Axis formatting | `percent_format()`, `comma_format()` |
| `effsize` | Effect sizes | `cohen.d()` |
| `coursepackR` | Course data and tools | `data(unified_music)`, `download_week()` |

## Statistical Test Selector

| Situation | Test | R Function |
|-----------|------|------------|
| Compare means of 2 groups | Two-sample t-test | `t.test(y ~ group, data = df)` |
| Compare means of 3+ groups | ANOVA | `aov(y ~ group, data = df)` |
| Two continuous variables | Pearson correlation | `cor.test(df$x, df$y)` |
| Two categorical variables | Chi-square | `chisq.test(table(df$x, df$y))` |

## Effect Size Quick Reference

| Test | Effect size | Function |
|------|-------------|----------|
| T-test | Cohen's d | `cohen.d(y ~ group, data = df)` |
| Correlation | r (already from `cor.test`) | — |
| Chi-square | Cramér's V | `sqrt(chi$statistic / (n * (k-1)))` |

## The Render → Push Cycle (Always in Terminal)

```bash
quarto render
git add .
git commit -m "Your message here"
git push
```

---

*MC 451 — Research Methods in Mass Media | Spring 2026 | SIUE Department of Mass Communications*
*Dr. Alex P. Leith | aleith@siue.edu*
