---
title: "Weekly Journal Entry"
subtitle: "Week `r params$week`: `r params$title`"
author: "Your Name"
date: today
params:
  week: 1
  title: "Untitled"
  chapters: "1"
  approach: "connector"
format:
  html:
    embed-resources: true
    toc: true
    theme: cosmo
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
---

```{r}
#| include: false

# Validate approach parameter
valid_approaches <- c("connector", "troubleshooter", "critic")
if (!tolower(params$approach) %in% valid_approaches) {
  stop("approach must be one of: ", paste(valid_approaches, collapse = ", "))
}
approach <- tolower(params$approach)
```

## Journal Metadata

| Field | Value |
|-------|-------|
| **Week** | `r params$week` |
| **Chapter(s) Read** | `r params$chapters` |
| **Approach** | `r tools::toTitleCase(params$approach)` |
| **Date** | `r Sys.Date()` |

---

## Instructions

Each week, reflect on course material using your chosen approach. Write 250--400 words of genuine reflection. Be honest---this is for *your* learning.

**To use this template:**

1. Copy this file to your `01_Journal/` folder
2. Rename it `Week_XX_Journal.qmd` (replacing XX with the week number)
3. Update the `params:` in the YAML header (week, title, chapters, approach)
4. Delete the two approaches you are NOT using
5. Write your reflection in the remaining section
6. Render to HTML or PDF in RStudio

---

`r if (approach == "connector") "## The Connector\n\nLink this week's material to something bigger: your project, a previous concept, real-world journalism, or your own experience.\n"`

`r if (approach == "connector") "**Connection to:**\n\n[What does this connect to? Your project? A previous concept? A real story?]\n\n**My Reflection:**\n\n[Write 250--400 words connecting this week's material to something meaningful...]\n\n**Link to my work:**\n\n- Related note, project, or earlier reflection\n"`

`r if (approach == "troubleshooter") "## The Troubleshooter\n\nDocument a challenge you faced and how you overcame it---a technical issue, confusing concept, or stuck analysis.\n"`

`r if (approach == "troubleshooter") "**Challenge I faced:**\n\n[What was difficult? An R error? A confusing concept? A stuck analysis?]\n\n**What I tried:**\n\n[What steps did you take to solve it?]\n\n**What worked:**\n\n[How did you overcome it? What did you learn?]\n\n**Resource that helped:**\n\n[Code snippet, chapter reference, online resource, etc.]\n"`

```{r}
#| eval: !expr approach == "troubleshooter"
#| echo: true
# Example code that helped:
# [Your code here]
```

`r if (approach == "troubleshooter") "**My Reflection:**\n\n[Write 250--400 words reflecting on the struggle and learning process...]\n"`

`r if (approach == "critic") "## The Critic\n\nAsk critical questions. Challenge an assumption, spot a gap in the material, or explore implications.\n"`

`r if (approach == "critic") "**Question I'm asking:**\n\n[What critical question arose from this material?]\n\n**What the material presented:**\n\n[Summarize the key idea you're interrogating.]\n\n**My critical take:**\n\n[What do I question about this? What gap exists? What implication concerns me?]\n\n**My Reflection:**\n\n[Write 250--400 words exploring this critical question deeply...]\n"`

---

*Rendered on `r Sys.time()` | MC-451 Research Methods | coursepackR*
